---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BOLD.NODE
**BOLD_NODE** is an R package that offers functionality to efficiently explore BOLD dataset releases (<https://boldsystems.org/data/data-packages/>) in the **Barcode Core Data Model (BCDM)** format locally. It uses a **DuckDB** backend to query **parquet** files directly in R, enabling fast searches even on systems with limited RAM. Data collection is optimized through customizable chunk sizes and configurable system pause intervals. 
The package also allows seamless conversion of search results into standard R data structures without collecting the data in memory for downstream analyses:

The package also allows seamless conversion of search results into standard R data structures without collecting the data in memory for downstream analyses:

1) **occurrence matrix** (biodiversity and ecology related analyses)
2) **sf** (spatial data analyses)
3) **DNAStringset** (phylogenetic data analyses)
4) **fasta** (phylogenetic data analyses in third-party tools or in R)

<!-- badges: start -->
<!-- badges: end -->

## Installation

The package can be installed using `devtools::install_github` function from the `devtools` package in R (which needs to be installed before)
```{r setup,warning=FALSE,message=F}



```

## Workflow

A typical workflow for exploring BOLD data (Steps in *italics* are optional but useful in some instances)
1) `bold.get.vocab` *(Provides unique terms present in a particular field, making it easier for exploring `bold.data.search` search parameters)*
2) `bold.data.search` (Searches the dataset based on the user criteria and prints the records in the search) 
3) `bold.concise.summary` *(Provides a detailed summary of the dataset retrieved)*
4) `bold.data.collect`(Collects the output of the `bold.data.search` in memory for downstream exploration/analyses).  
Details on the BCDM fields (field names and definitions) can be obtained using the `bold.fields.info()` function or on the BCDM GitHub page (<https://github.com/boldsystems-central/BCDM>).


### Get the vocabulary for specific fields

```{r,echo=FALSE}

# parquet_file<-'path where the parquet file from BOLD is downloaded'

# vocab.data <- bold.get.vocab(parquet_file,specific.cols = c("country/ocean"))

```

### Search the dataset

```{r,echo=FALSE}

# parquet_file<-'path where the parquet file from BOLD is downloaded'

# Taxonomy 
# bold_search_taxonomy <- bold.data.search(parquet_path=parquet_file,
# taxonomy = c("Odonata","Poecilia"))

# Geography
# bold_search_geography <- bold.data.search(parquet_path=parquet_file,
# taxonomy = c("Panthera pardus),
# geography = c("India"))

# Combination of many search criteria
# bold_search_combination <- bold.data.search(
# parquet_path=parquet_file,
# taxonomy = "Coleoptera",
# geography = "Canada",
# marker = "COI-5P",
# basecount = c(500, 660)

```

### Data summary

```{r,echo=FALSE}

# Get the concise summary

# bold_summary <- get.concise.summary(bold_search_geography)

```

### collect the search data

```{r,echo=F}

# Collect data (no export)
# collected_data<-bold.data.collect(
# bold_search_geography,
# chunk.size = 50000,
# export = FALSE)

# Collect data (with parquet export)

# bold.data.collect(
# bold_search_combination,
# chunk.size = 50000,
# export = TRUE,
# export.type = "parquet",
# output.path = userdefinedpath)


```

**Please note** Some queries (e.g., All "Diptera") may return very large datasets. Always check the summary before collecting data to ensure you don't exceed the available RAM.
