---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BOLD.NODE

**BOLD_NODE** is an R package that offers functionality to efficiently explore BOLD dataset releases (<https://boldsystems.org/data/data-packages/>) in the **Barcode Core Data Model (BCDM)** format (for more information on BCDM please visit its GitHub repo <https://github.com/boldsystems-central/BCDM>) locally. It uses a **DuckDB** back end to query **parquet** files directly in R, enabling fast searches even on systems with limited RAM. Data collection is optimized through customized chunk sizes and configurable system pause intervals. 

The package also allows seamless conversion of search results into standard R data structures without collecting the data in memory for downstream analyses:

1. **occurrence matrix** (biodiversity and ecology related analyses)
2. **sf** (spatial data analyses)
3. **DNAStringset** (phylogenetic data analyses)
4. **fasta** (phylogenetic data analyses in third-party tools or in R)

<!-- badges: start -->
<!-- badges: end -->

## Installation

The package can be installed using `devtools::install_github` function from the `devtools` package in R (which needs to be installed before)
```{r setup,warning=FALSE,message=F,results='hide'}

devtools::install_github('https://github.com/sameerpadhye/BOLD.NODE.git')

```


## Downloading Data Packages

Users need to log into BOLD (<https://bench.boldsystems.org/index.php/Login/page?destination=MAS_Management_UserConsole>) to download the datasets in the `parquet` format. The users can then directly use the file as an input for the search and vocabulary functions


## BOLD.NODE has 10 functions:

1. bold.bcdm.fields
2. bold.data.search
3. bold.data.collect
4. get.concise.summary
5. *get.DNAStringset*
6. get.DwC
7. get.fasta
8. get.occ.data
9. get.sf
10. get.vocab


**Note**
*Function 5*: *get.DNAStringset* requires the package `Biostrings` to be installed and imported in the R session beforehand. It can be installed using using `BiocManager` package.


```{r Biostrings install,warning=FALSE,message=FALSE}

# if (!requireNamespace("BiocManager", quietly=TRUE))
#   
# install.packages("BiocManager")
# 
# BiocManager::install("Biostrings")

```

## Workflow for search and collect 

A typical workflow for exploring and  BOLD data (Steps in *italics* are optional but useful in some instances):

1. `bold.get.vocab` *(Provides unique terms present in a particular field, making it easier for exploring* `bold.data.search` *search parameters)*
2. `bold.data.search` (Searches the dataset based on the user criteria and prints the number of records available) 
3. `bold.data.collect`(Collects the output of the `bold.data.search` in memory for downstream exploration/analyses).  

### 1.Get the vocabulary for specific fields 
This function can be used for getting unique values of some of the categorical fields (e.g. institutes) to make searches easier

```{r get_vocab}

# parquet_file<-'path where the parquet file from BOLD is downloaded'

# vocab.data <- bold.get.vocab(parquet_file,specific.cols = c("country.ocean"))

```

### 2.Search the dataset
This function lets users search by more than 10 different search parameters including taxonomy (species, genus, family etc.), geography (country.ocean, province, region), ids (processid, sampleid) and more. The users can just input the search query directly (e.g. put *Canada* as a query in the `geography` argument or *Coleoptera* in the `taxonomy` argument) with the function searching the query in the relevant fields internally. The search result is a `tbl_sql` object that is used for collecting or transforming the data.

```{r data_search}

# parquet_file<-'path where the parquet file from BOLD is downloaded'

#1 Taxonomy 

# bold_search_taxonomy <- bold.data.search(parquet_path=parquet_file,
# taxonomy = c("Odonata","Poecilia"))

#2 Geography

# bold_search_geography <- bold.data.search(parquet_path=parquet_file,
# taxonomy = c("Panthera pardus),
# geography = c("India"))

#3 Combination of many search criteria

# bold_search_combination <- bold.data.search(
# parquet_path=parquet_file,
# taxonomy = "Coleoptera",
# geography = "Canada",
# marker = "COI-5P",
# basecount = c(500, 660)

```

### 3.Collect the searched data
The searched data can be collected in memory using this function. **Please note** Some queries (e.g., All "Diptera") may return very large datasets. Always check the printed message in the console (shows the total records in the search) after search before collecting data to ensure you don't exceed the available RAM.

```{r data_collect}

# Collect data (no export)

# collected_data<-bold.data.collect(
# bold_search_geography,
# chunk.size = 50000,
# export = FALSE)

# Collect data (with parquet export)

# bold.data.collect(
# bold_search_combination,
# chunk.size = 50000,
# export = TRUE,
# export.type = "parquet",
# output.path = userdefinedpath)


```


### The `get.` functionality
The `get.` functions convert the search results from the `bold.data.search` into objects used in packages such as `vegan`, `msa`, `DECIPHER`, `terra`, `geodata` etc.


#### `get.concise.summary` 
gets a concise summary of the searched data. Search results include total records, total countries, amplicon length range and many more

```{r get.concise.summary,warning=F}

#  Search the data

# bold_search <- bold.data.search(
# input.parquet=parquet_file,
# taxonomy = "Coleoptera",
# geography = "Canada",
# marker = "COI-5P",
# basecount = c(500, 660))


# Get concise summary of the data

# bold_summary <- get.concise.summary(bold_search)

```

#### `get.fasta` 
generates a custom header fasta file of the searched data. This can be exported for any downstream analytical pipelines

```{r get.fasta,warning=F} 

#  Search the data

# bold_search <- bold.data.search(
# input.parquet=parquet_file,
# taxonomy = "Coleoptera",
# geography = "Canada",
# marker = "COI-5P",
# basecount = c(500, 660))

# Get fasta

# get.fasta(
# bold_search,
# output.file = "trial.fas",
# fas.header = c("bin_uri", "processid"))

```

#### `get.sf`
generates a `sf` object of the searched data for any downstream spatial data analyses

```{r get concise_summary,warning=F} 

#  Search the data

# bold_search <- bold.data.search(
# input.parquet=parquet_file,
# taxonomy = "Coleoptera",
# geography = "Canada",
# marker = "COI-5P",
# basecount = c(500, 660))

# Get sf

# sf_res<- get.sf(bold_search, chunk.size = 100000)

```

#### `get.occ.data`
creates an occurrence matrix from the searched data based on the `taxon.rank`, `taxon.name` (optional) and the `site.cat`

```{r get.occ.data,warning=F} 

#  Search the data

# bold_search <- bold.data.search(
# input.parquet=parquet_file,
# taxonomy = "Coleoptera",
# geography = "Canada",
# marker = "COI-5P",
# basecount = c(500, 660))

# Get occurrence data

# occurrence_data <- get.occ.data(
#   bold_search,
#   taxon.rank = "family",
#   site.cat = "region")

```

#### `get.DNAStringSet` 
generates a `DNAStringSet` (Biostrings object) object of the searched data for any downstream sequence alignment and tree generation with custom headers. The library `Biostrings` has to be installed and imported before using this function

```{r get.DNAStringSet,warning=F} 

#  Search the data

# bold_search <- bold.data.search(
# input.parquet=parquet_file,
# taxonomy = "Coleoptera",
# geography = "Canada",
# marker = "COI-5P",
# basecount = c(500, 660))

# Get DNAStringSet

# bold.dnastringset<-get.DNAStringSet(bold_search,
# marker="COI-5P",
# cols_for_seq_names = c("processid","family"))
```


```{r benchmarks_data,include=F,warning=F}

library(ggplot2)
benchmarks<-read.csv("G:/Research data/CBG_work/APIless_BOLD_datarelease_search/benchmarks_new.csv")
benchmarks$Category<-factor(benchmarks$Category,levels=c("<1k records", "1k-100k records","100k-1M records","1M-5M records",">5M"))

```

#### It takes roughly 10 minutes to collect ~10M records on a i7 2.8GHZ 16GB RAM machine
```{r benchmark_fig,fig.width=6, fig.height=4, dpi=300, dev='jpeg',echo=F,warning=F,message=F}

ggplot(benchmarks,
       aes(x=Category,
             y=median_time_collect))+
  geom_line(linewidth = 1.3, 
            color = "black") +
  geom_point(
    size = 5,
    pch = 21,
    fill = 'orangered',
    color = "black",
    alpha = 0.7
  ) +
  geom_text(aes(label = exact_records), vjust = 0.3, hjust = 1.5)+
  labs(
    x = "",
    y = "Download time in secs",
    title = "Benchmarks: Data collection (download) with BOLD.NODE") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()) 

```

*The package is under active development and  functionality is subject to change*
