---
title: "Using_BOLD.NODE"
author: "Sameer Padhye, Spencer K. Monckton, Dirk Stienke, Jireh Agda, Terresita Porter"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    highlight: pygments
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
    fig_width: 8
    fig_height: 5
    self_contained: true
vignette: >
  %\VignetteIndexEntry{basic_workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{=html}
<style>
body {
  font-size: 18px;
  line-height: 1.7;
}

h1 {
  font-weight: 600;
  color: #2c3e50;
  border-bottom: 2px solid #3498db;
  padding-bottom: 0.4em;
}

h2 {
  font-weight: 600;
  color: #34495e;
  margin-top: 2em;
}

h3 {
  color: #2980b9;
}

pre {
  background: #f6f8fa;
  border: 1px solid #e1e4e8;
  border-radius: 6px;
  padding: 1em;
}

code {
  background: #f1f3f5;
  color: #c7254e;
  padding: 0.15em 0.4em;
  border-radius: 3px;
  font-size: 0.88em;
}

pre code {
  background: transparent;
  color: inherit;
  padding: 0;
}

.note {
  background: #e8f4fd;
  border-left: 4px solid #3498db;
  padding: 1em 1.2em;
  border-radius: 0 6px 6px 0;
  margin: 1.5em 0;
}

.warning {
  background: #fef9e7;
  border-left: 4px solid #f39c12;
  padding: 1em 1.2em;
  border-radius: 0 6px 6px 0;
  margin: 1.5em 0;
}

table {
  border-collapse: separate;
  border-spacing: 0;
  border-radius: 6px;
  overflow: hidden;
  box-shadow: 0 1px 4px rgba(0,0,0,0.06);
}

thead th {
  background: #3498db;
  color: white;
  border: none;
  padding: 0.8em 1em;
  font-size: 16px;
  line-height: 1.2;
}

tbody tr:nth-child(even) {
  background: #f8f9fa;
}

tbody td {
  padding: 0.7em 1em;
  border-top: 1px solid #ecf0f1;
}

#TOC {
  border-right: 1px solid #ecf0f1;
  padding-right: 1em;
}

.dataTables_wrapper {
  max-width: 1000px;
  margin: 0 auto;
}

table.dataTable {
  width: 100% !important;
}

div.dataTables_scrollBody {
  overflow-x: auto !important;
}
</style>
```

# Introduction

This vignette demonstrates a complete workflow integrating the *search*, *collect*, *get_* and *bcdm_to* functions in `BOLD.NODE` (No API Offline Data Explorer) for efficient and scalable exploration of BOLD data packages. The latest data package parquet file can be downloaded from the BOLD website (<https://boldsystems.org/data/data-packages/>). As a case study, we use Canadian **Cerambycidae** (longhorn beetles) to illustrate how these functions can be combined to search, retrieve, and transform biodiversity data. We also demonstrate how outputs generated by the *bcdm_to* functions can serve as inputs for downstream analyses in other R packages, enabling integration with broader biodiversity and ecological workflows.

# Objectives

1.  Search all public Canadian Cerambycidae records from the BOLD data package.
2.  Collect (download) the search.
3.  Summarize the retrieved data using a concise overview.
4.  Determine the lowest supported taxonomic identification for each Cerambycidae BIN through BIN-level taxonomic concordance.
5.  Construct a `BIN × province` occurrence matrix for downstream biodiversity analyses with the vegan package.
6.  Obtain representative records for each BIN based on sequence length and identification method.
7.  Extract *Clytus* sequences as a `DNAStringSet` object and calculate nucleotide base proportions using the `Biostrings` package.
8.  Convert *Monochamus* occurrence records into an `sf` object and visualize their geographic distribution.

# Practical Workflow

## Installation of the package

The latest (in development) version of BOLD.NODE can be installed via `devtools`

```{r install,message=F,warning=F,echo=FALSE}
# install.packages("devtools")
# Install BOLD.NODE (offline Parquet explorer)
# devtools::install_github("https://github.com/sameerpadhye/BOLD.NODE.git")
library(BOLD.NODE)
```

The `bcdm_to_dnastringset` function from the package requires `Biostrings` to be installed before use.

```{r install-packages,warning=F,message=F}
# Install Bioconductor dependencies for sequence alignment
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install("Biostrings")

# Load Biostrings
library(Biostrings)
```

Some additional libraries like `dplyr`, `vegan` are imported for the workflow as well.

```{r dplyr-install,include=T}
library(dplyr)
library(ggplot2)
library(sf)
library(vegan)
```

## Data package import

The path where the file is downloaded is saved as a variable that can be used as the input for the `search` function.

```{r setup-parquet_toshow}
# Path to the downloaded BOLD Parquet release file
# parquet_file <- "G:/usr/path/file.parquet"
```

```{r setup-parquet,include=F}
# Path to the downloaded BOLD Parquet release file
parquet_file <- '/Users/sameerpadhye/Desktop/R data/Data/BOLD_Public.03-Jul-2026.parquet'
```

### 1. Search from the data package

`bold_parquet_search` queries the parquet data package using user-defined search terms such as **COI-5P** for marker type, **Cerambycidae** for taxonomic group, and **Canada** for geographic region.

```{r search,message=F,warning=F}
# Search the BOLD dataset
cerambycidae_search <- bold_parquet_search(
  input.parquet = parquet_file,
  taxonomy = "Cerambycidae",
  geography = "Canada",
  marker = "COI-5P"
)
```

```{r search-records,include=F}
tot_records <- cerambycidae_search %>%
  dplyr::summarise(Total_records = dplyr::n()) %>%
  dplyr::collect() %>%
  dplyr::pull(Total_records)
```

The total number of records is shown in the console.

```{r print_records,echo=FALSE}
cat("The search has", tot_records, "records in the dataset")
```

### 2. Collect the dataset

The search result is then collected into memory (i.e., downloaded in the R session) as a data frame using `bold_search_collect` (*Please note that only first 100 results and a few columns are shown here*).

```{r collect,warning=F,message=F}
# Collect the search results into memory
cerambycidae_data <- bold_search_collect(
  cerambycidae_search,
  chunk.size = 50000,
  export = FALSE
)
```

```{r display-tabular_output,message=F,warning=F}
# Inspect the collected data
DT::datatable(
  head(cerambycidae_data, 100) %>% select(processid, sampleid, bin_uri, family, genus, species),
  options = list(pageLength = 10, scrollX = TRUE)
)
```

### 3. Generate a concise summary

`get_concise_summary` provides a summary of the search results, including total records, unique species, BINs, countries, institutes, markers, and marker length range.

```{r summary,message=F,warning=F}
# Generate concise summary
cerambycidae_summary <- get_concise_summary(cerambycidae_search)
DT::datatable(cerambycidae_summary)
```

### 4. Compute BIN consensus taxonomy

`get_bin_consensus` computes the consensus taxonomic identifications for each BIN in the Cerambycidae dataset. The function works upwards through taxonomic ranks to find the lowest concordant identification that satisfies the provided threshold (*Please note that only first 50 results are shown here*).

```{r step4-bin-consensus,message=F,warning=F}
# Get BIN consensus with strict consensus threshold (1.0)
bin_consensus <- get_bin_consensus(
  cerambycidae_search,
  threshold = 1.0,
  min.ids = 1
)

# View consensus at different taxonomic ranks
table(bin_consensus$concordant_rank)

# Examine the results
DT::datatable(head(bin_consensus, 50),
  options = list(
    pageLength = 10,
    scrollX = TRUE
  )
)
```

### 5. Obtain a BIN-by-province occurrence matrix for all cerambycids

`bcdm_to_occmatrix` creates a `site-by-taxon` occurrence matrix from the full Cerambycidae dataset, using **province.state** as the geographic grouping variable and **species** as the taxonomic rank (*Please note only 20 species are shown here*).

```{r step5-occurrence-matrix,message=F,warning=F}
# Generate occurrence matrix at species level, grouped by province/state
occ_matrix <- bcdm_to_occmatrix(
  cerambycidae_search,
  kingdom = "Animalia",
  taxon.rank = "species",
  site.cat = "province.state"
)
DT::datatable(
  head(occ_matrix, 20),
  options = list(
    pageLength = 10,
    scrollX = TRUE
  )
)
```

The resulting occurrence matrix can be directly used as input data for many functions in widely used packages like `vegan`.

```{r install-vegan, include=F,message=F,warning=F}
library(vegan)
```

```{r vegan-example,message=F,warning=F}
species_richness_est <- poolaccum(occ_matrix)
species_richness_est
```

### 6. Select representative records for each BIN

`get_bin_reps` sub-samples the search results by BIN and selects representative records according to the provided criteria, yielding a representative sample for downstream analyses. As we saw above, some BINs include multiple species, therefore we will select representatives for all BIN-taxon combinations. Here, we select up to three records per BIN, prioritizing 658-bp sequences from specimens identified via morphological examination.

```{r step5-bin-reps,message=F,warning=F}
# Sample 3 records per BIN-taxon combination
bin_reps <- get_bin_reps(
  bold.search.res = cerambycidae_search,
  Nreps = 3,
  by.tax = TRUE,
  enforce.scientific = TRUE,
  criteria = list(seq_length = 658,
                  id_method = "Morphology",
                  vouchered = TRUE)
)

# Compare sample to full dataset
n_bins <- length(unique(bin_reps$bin_uri))
cat("Sampled", nrow(bin_reps), "representatives from", n_bins, "BINs",
    "(out of", tot_records, "total records).")
```

The actual number of sampled records varies according to BIN size and taxonomy, as we can see in the collected results (*Please note only 50 records are shown here*). In the case of BINs with multiple species, further analysis of barcode sequences often reveals small but consistent differences between them.

```{r step5-bin-rep-DT,message=F,warning=F}
DT::datatable(
  head(bin_reps, 50),
  options = list(
    pageLength = 10,
    scrollX = TRUE
  )
)
```

### 7. Generate a DNAStringset object for the genus *Clytus*

`bcdm_to_dnastringset` converts the search results into a `DNAStringSet` object. First, `bold_parquet_search` is used to filter the COI-5P *Clytus* records from Canada. Additional search parameters are put in to return only full length barcodes having very low (or none) ambiguous bases (**Only top three results displayed**).

```{r import-Biostrings, include=F}
library(Biostrings)
```

```{r DNAStringSet-object,message=F,warning=F}
# Filter for genus Clytus
clytus_search <- bold_parquet_search(
  input.parquet = parquet_file,
  taxonomy = "Clytus",
  geography = "Canada",
  marker = "COI-5P",
  basecount = 658,
  ambi.base.cutoff = "<1%"
)

# Collect the filtered data
clytus_data <- bcdm_to_dnastringset(clytus_search,
  cols_for_seq_names = c("processid", "bin_uri")
)
head(clytus_data, 3)
```

The `DNAStringSet` object can then be used in various downstream analysis using packages like `Biostrings`. Here we specifically look at the nucleotide base proportions in the COI-5P sequences of the genus `Clytus`.

```{r basepair-freq,message=F,warning=F}
# Count bases across all sequences
base_counts <- colSums(alphabetFrequency(clytus_data, baseOnly = TRUE))
# Proportions
base_props <- base_counts / sum(base_counts)
base_props
```

### 8. Generate a `sf` object for spatial mapping of occurrences of the genus *Monochamus*

```{r sf_object}
# Filter for genus Monochamus
monochamus_search <- bold_parquet_search(
  input.parquet = parquet_file,
  taxonomy = "Monochamus",
  geography = "Canada",
  marker = "COI-5P"
)

# Collect the filtered data
monochamus_data <- bcdm_to_sf(monochamus_search)
```

```{r ggplot-import,include=F}
library(ggplot2)
library(sf)
```

The `sf` object can be used directly for mapping occurrences with packages like `ggplot2`.

```{r occurrence-map}
# Creating a background map using the maps package; some map_data country names (ID column) are changed to suit the BCDM country.ocean names
map_data <- sf::st_as_sf(maps::map("world",
  plot = FALSE,
  fill = TRUE
)) %>%
  filter(ID == "Canada")
# Convert the data to WGS84
map_data <- st_transform(
  map_data,
  4326
)
# Plot
map_plot <- ggplot() +
  geom_sf(
    data = map_data,
    alpha = 0.3,
    linewidth = 0.4
  ) +
  geom_point(
    data = monochamus_data,
    mapping = aes(
      x = lon,
      y = lat
    ),
    colour = "#011B26",
    fill = "#F78E1E",
    size = 3,
    pch = 21
  ) +
  theme_bw(base_size = 15) +
  theme(panel.grid.major = element_line(
    colour = "grey50",
    size = 0.3,
    linetype = 3
  )) +
  xlab("Longitude") +
  ylab("Latitude") +
  coord_sf(expand = FALSE) +
  ggtitle("Distribution map")

map_plot
```

# Performance and Integration Notes

`BOLD.NODE` can search and collect very large data very effectively (the collect part is dependent on the user's machine, however, as very large searches can overload available memory). The functions from this package can also be combined with the analysis functions from the `BOLDconnectR` package directly due to the underlying `BCDM` data model (<https://github.com/boldsystems-central/BCDM>). For the `bold_parquet_search` and `bold_search_collect` function benchmarks, refer to <https://github.com/sameerpadhye/BOLD.NODE> for details.
